\Title

### 2022-10-28 Grimaldi-etal-BiolCybernetics.tex
Update on Overleaf.

@@

%\ninept
%
%\abstract{Abstract}
%
\keywords{\Keywords}
%
\maketitle
%%%%%%%%% BODY TEXT
\section{Introduction}
%%%-----------------------------------------------------------------
%%%-----------------------------------------------------------------
\label{sec:intro}
% >>>>>>>>>>> LuP is here <<<<<<<<<<<
%
\subsection{Recent challenges in machine learning and computer vision}
%%%-----------------------------------------------------------------
Machine learning, and computer vision in particular, have undergone a major transformation in modern times, transitioning from an active character in science fiction movies to become mainstream tools found in the smartphones in everyone's pocket. Less visible, yet crucial to our economy and security, these automated tools now dominate the range of services that equip virtually most sectors of our daily lives. This trend is profoundly pronounced in the field of medicine, as it provides practitioners with new diagnostic tools to treat people faster and more effectively. This particular example also illustrates some important challenges that these tools are increasingly facing. One of these is the sensitivity of deep-learning based algorithms to adversarial attacks, that is, the sensitivity of the models' output to minute changes in the input. This could raise important questions about the reliability of the results of any automated diagnostic model. More generally, these models do give an output for any possible input but they lack an explanation as to why they produced it. This lack of explainability is a hindrance to the advancement of these methodologies as well as to their acceptance for most safety-relevant tasks. Having said that, we may remember that these tools are inherited from theoretical models originally designed to understand the brain. While the first artificial neural networks were mainly applied to toy-like problems, they also offered a growing number of applications in machine learning and computer vision, but also more generally for task automation. In particular, convolutional neural networks are inspired by the feed-forward processing observed in the visual pathways of primates, and today they reach higher performances than humans for key image categorization challenges. One of the challenges of these models is to be able to scale them up for future applications, for example in autonomous driving. Indeed, in such applications, the input does not consist of a single image (say a few million pixels) but of a continuous stream of images from several cameras.

The challenges are numerous in such situations and two major obstacles must be highlighted: the computational delay and the energy budget. First, most existing solutions are built incrementally from models built to process low definition images, such as Le Net for MNIST~\citep{lecun_gradient-based_1998}, then high definition images, such as VGG on Imagenet~\citep{simonyan_very_2015}. This has a cost and notably the fact that the computation time increases with the size of the image. This phenomenon has been mitigated by the increase in available computing power, notably thanks to the parallel processing implemented in GPU cards. However, the transition from images to videos, then to multiple cameras implies a further level in the change of scale in the amount of data processed. It also raises an issue that is often not addressed when it comes to image categorization. Indeed, in autonomous driving, an important task is to react as quickly as possible to a possible collision, for example when a pedestrian unexpectedly crosses the street. As such, future systems should always be able to give a response at any time, instead of having to wait for the pipeline to provide the correct answer. This online processing is also a feature of processing in biological systems and could be an important first contribution if we could translate these principles into future automated processes. %
Second, a major constraint for future systems is energy cost. As we have seen, fast computation time has been maintained for high-dimensional images, but at the cost of higher energy consumption. In addition, the need to solve increasingly complex tasks has required the construction of increasingly complex and deep architectures. As an example, the best performing model in the Imagenet challenge uses 2100 million parameters. Training the model typically consumes kilowatts of power and about 1 joule to infer a decision per single image. Less power-hungry solutions have emerged, such as Mobile Net~\citep{howard_searching_2019}, but they are not yet ready for practical use in autonomous cars, as the additional power consumption would be a significant percentage of the energy consumed by an electric car under standard conditions.

@@

Neuroscience could provide revolutionary answers to these two closely related problems. Indeed, the human brain has the remarkable property of being able to react at any time, while consuming a reasonable amount of energy, about tens of watts. This system is the result of millions of years of natural selection, and a striking difference between biological neural networks and artificial ones is the representation they use. CNNs, for example, represent information passing from one layer to another as tensors representing visual information densely over the visual topography, while representing different properties in different channels. These networks have been known to mimic many properties of biological systems~\citep{schrimpf_brain-score_2020}, yet, this score does not count rapidity or energy. On the other hand, information in a large majority of biological neural networks is represented as spikes, i.e. prototypical all-or-nothing events whose only parameters are their timing and the address of the neuron that emitted this spike~\citep{paugam-moisy_computing_2012}. Why spikes? It is not yet known why most biological systems have evolved to use this representation, but one major advantage seems to be some form of efficiency. In the human brain, each of the approximately 90 trillion nerve cells emits on average one spike per second which fan out to an average of 10000 synapses. Importantly, each neuron fires with great variability, and up to 300 spikes per second. Many attempts have been made to build artificial spiking neural networks (SNNs), and in particular to achieve the same performance as observed in classical CNNs with a lower energy budget. However, none of these networks has yet surpassed the state of the art performances. Here, we will try to focus on one core property of spiking neurons inspired by neurobiology to progress in this direction.

@@

Let us first review some existing SNN models. A first motivation for these models was to understand biological observations at the different scales of the central nervous system. For example, the Hodgkin-Huxley model describes the dynamics of the membrane potential of a single nerve cell and specifically explains the role of the different voltage-dependent ion channels. It also explains how the model can produce a spike. Larger scale models introduce an simpler dynamics and an explicit spiking mechanism, for example in the Integrate-and-Fire model. It predicts the emergent states of a recurrently connected neural network~\citep{brunel_phase_2000}. Several models have increased the scale of the system to include more neurons, such as The Virtual Brain (TVB)~\citep{sanz_leon_virtual_2013}, which mixes different scales and includes SNNs with (analog) neural mass models. However, these methods are mainly descriptive with applications in neuroscience but few spiking neural models have had an application in computer vision. Another class of approaches have designed normative models of SNNs that aim to have applications in machine learning. One of these is the SpikeNet algorithm which uses a purely temporal approach by encoding information using one spike per neuron~\citep{delorme_spikenet_1999}. Another type of SNN using precise spike timing attempts to determine the structure of the network in order to minimize a cost function. This was implemented in the SpikeProp algorithm~\citep{bohte_error-backpropagation_2002} and has been extanded in novel gradient-based methods. The surrogate gradient method is now widely used in methods that attempt to transfer performance from CNNs to SNNs~\citep{zenke_remarkable_2021}. However, the performance of SNNs is still lagging that of analog networks and the question of the advantage of using spikes in machine learning and computer vision remains open.

@@

Let us highlight some innovations in the contributions presented in this paper. First, the generic heterogeneous model is formalized from first principles for optimal detection of the event-based spatio-temporal motifs, whereas~\citet{ghosh_spatiotemporal_2019,yu_stsc-snn_2022} use a correlation-based heuristic, which we have observed to be less efficient. Moreover, in comparison to HOTS~\citep{lagorce_hots_2017} the weights are explainable as they directly inform on the logit (inverse sigmoid of the probability) of detecting each spatio-temporal spiking motif. Another novelty is that the model simultaneously learns the weights and the delays, while for example the polychronization model~\citep{izhikevich_polychronization_2006} only learns the weights using STDP while the delays are randomly drawn and their values frozen. Moreover, the model is evaluated on a realistic task, while models such as the tempotron are tested on simplified problems~\citep{gutig_tempotron_2006}. Another main contribution is to provide a model that is suitable for learning any type of spatio-temporal spiking motifs and that can be trained in a supervised way by providing a dataset of supervision pairs. This allows for a more flexible definition of the model using this properly labeled dataset instead of relying on a careful description of the physical rules governing a task, e.g., the luminance conservation principle for motion detection~\citep{benosman_asynchronous_2012, dardelet_event-by-event_2021}.
%
\subsection{Main limits}
%%%-----------------------------------------------------------------
We have identified several limits to our model which we will now detail. First, the complete framework is based on a discrete binning of time that is incompatible with the continuous nature of biological time. We have used this binning to be able to efficiently implement the framework on conventional hardware, especially GPUs, and in particular to be able to use three-dimensional convolutions. We have tested the effect of the size of the time bin and shown that it has essentially no impact on the results presented in this paper. This is consistent with the relative robustness of other event-based frameworks such as HOTS~\citep{lagorce_hots_2017}, where the accuracy was not affected when the input spikes were subjected to noisy perturbations up to $1~\ms$~\citep{grimaldi_robust_2022}. This suggests the possibility of analytically including a precision term in the temporal value of the input spikes, a mechanism potentially implemented by the filtering that is implemented by the synaptic time constant of about $5~\ms$. Furthermore, it is possible to circonvent the necessity of using a time discretization by using a purely event-based scheme. Indeed, it is not necessary to compute voltage traces between two spikes~\citep{hanuschkin_general_2010} and it is thus possible to define a purely event-based framework. Such an architecture could provide promising speed gains for the calculations.  \note{pourrait parler du poster bernstein? de nadafian?}

Another restriction is that this model is purely feed-forward. Therefore, the spikes generated by the postsynaptic neurons are produced solely on the basis of information contained in the classical receptive field. However, it is well known that neurons in a layer can interact using lateral interactions, for instance in V1 and that this can be the basis for computational principles~\citep{chavane_revisiting_2022}. For example, the combination of neighboring orientations may contribute to image categorization~\citep{perrinet_edge_2015}. Furthermore, neural information is modulated by feedback information, for example to distinguish a figure from its background~\citep{roelfsema_early_2016} and it has been shown that feedback may be essential for building realistic models of primary visual areas~\citep{boutin_sparse_2020, boutin_effect_2020}, Notably to explain non-linear mechanisms~\citep{boutin_pooling_2022}. It is currently not possible to implement these recurrent connections in our implementation (lateral or feedback), mainly due to our use of convolutions. However, the generic theoretical model is able to include them by inserting new spikes into the list of spikes that are reaching pre-synaptic addresses. While this is possible in theory, it needs to be properly adapted in practice so that these recurrent connections do not amplify neuronal activity outside of a homeostatic state (through a extinction or explosion of activity).  
\note{ dire que la prédiction permet de faire de la prediction et que ça explique les résultats de leBec}%~\citep{perrinet_motion-based_2012}

Such recurrent activity would be essential for implementing predictive or anticipatory processes. This is essential in a neural system as it contains multiple different delays that require temporal alignment~\citep{hogendoorn_predictive_2019}. This has been previously modeled to explain for example the flash-lag illusion~\citep{khoei_flash-lag_2017}. \note{Say that this could be implemented by properly setting the timing of the label} As previously stated, this could be implemented using generalized coordinates (that is variables such as position which are complemented by speed, acceleration, jerk, ...) and ``neurobiologically, the application of delay operators just means changing synaptic connection strengths to take different mixtures of generalized sensations and their prediction errors''~\citep{perrinet_active_2014}. Our proposed model using heterogeneous delays provides an alternate and elegant implementation solution to this problem.
%
\subsection{Perspectives}
%%%-----------------------------------------------------------------
\note{la tache ne met pas en avant les capacites des motifs hetero synaptiques =trop simple et peut se faire en HOTS}
In the definition of our task, we highlighted how the generation of events depends on the spatial gradient in each image. This gradient is defined in both horizontal and vertical dimensions and its maxima are generally oriented. Taken together, these oriented edges form the contours of visual objects in the scene~\citep{koenderink_representation_1987}. Thus, there is a dependence within the event stream between motion and orientation information. It would be crucial to study this interdependence further. This could be initiated by training the model on a dataset with labels providing local orientation. \note{we can say we tried and that it works well}. Studying this dependence will allow us to dissociate these two forms of visual information and to enable us to integrate them. In particular it will allow us to consider that the definition of motion is more accurate perpendicular to an oriented contour (aka, the aperture problem), thus allowing us to implement recurrent predictive rules, such as those identified to dissociate this problem~\citep{perrinet_motion-based_2012}.

>>>>

% https://www.springer.com/journal/422/updates/20421374

@@

%\ninept
%
%\abstract{Abstract}
%
\keywords{\Keywords}
%
\maketitle
%%%%%%%%% BODY TEXT
\section{Introduction}
%%%-----------------------------------------------------------------
%%%-----------------------------------------------------------------
\label{sec:intro}
% >>>>>>>>>>> LuP is here <<<<<<<<<<<
%
\subsection{Recent challenges in machine learning and computer vision}
%%%-----------------------------------------------------------------
Machine learning, and computer vision in particular, have undergone a major transformation in modern times, transitioning from an active character in science fiction movies to become mainstream tools found in the smartphones in everyone's pocket. Less visible, yet crucial to our economy and security, these automated tools now dominate the range of services that equip virtually most sectors of our daily lives. This trend is profoundly pronounced in the field of medicine, as it provides practitioners with new diagnostic tools to treat people faster and more effectively. This particular example also illustrates some important challenges that these tools are increasingly facing. One of these is the sensitivity of deep-learning based algorithms to adversarial attacks, that is, the sensitivity of the models' output to minute changes in the input. This could raise important questions about the reliability of the results of any automated diagnostic model. More generally, these models do give an output for any possible input, but they lack an explanation as to why they produced it. This lack of explainability is a hindrance to the advancement of these methodologies, as well as to their acceptance for most safety-relevant tasks. Having said that, we may remember that these tools are inherited from theoretical models originally designed to understand the brain. While the first artificial neural networks were mainly applied to toy-like problems, they also offered a growing number of applications in machine learning and computer vision, but also more generally for task automation. In particular, convolutional neural networks are inspired by the feed-forward processing observed in the visual pathways of primates, and today they reach higher performances than humans for key image categorization challenges. One of the challenges of these models is to be able to scale them up for future applications, for example in autonomous driving. Indeed, in such applications, the input does not consist of a single image (say a few million pixels) but of a continuous stream of images from several cameras.

The challenges are numerous in such situations, and two major obstacles must be highlighted: the computational delay and the energy budget. First, most existing solutions are built incrementally from models built to process low definition images, such as Le Net for MNIST~\citep{lecun_gradient-based_1998}, then high definition images, such as VGG on Imagenet~\citep{simonyan_very_2015}. This has a cost and notably the fact that the computation time increases with the size of the image. This phenomenon has been mitigated by the increase in available computing power, notably thanks to the parallel processing implemented in GPU cards. However, the transition from images to videos, then to multiple cameras implies a further level in the change of scale in the amount of data processed. It also raises an issue that is often not addressed when it comes to image categorization. Indeed, in autonomous driving, an important task is to react as quickly as possible to a possible collision, for example when a pedestrian unexpectedly crosses the street. As such, future systems should always be able to give a response at any time, instead of having to wait for the pipeline to provide the correct answer. This online processing is also a feature of processing in biological systems and could be an important first contribution if we could translate these principles into future automated processes. %
Second, a major constraint for future systems is energy cost. As we have seen, fast computation time has been maintained for high-dimensional images, but at the cost of higher energy consumption. In addition, the need to solve increasingly complex tasks has required the construction of increasingly complex and deep architectures. As an example, the best performing model in the Imagenet challenge uses 2100 million parameters. Training the model typically consumes kilowatts of power and about 1 joule to infer a decision per single image. Less power-hungry solutions have emerged, such as Mobile Net~\citep{howard_searching_2019}, but they are not yet ready for practical use in autonomous cars, as the additional power consumption would be a significant percentage of the energy consumed by an electric car under standard conditions.

@@

Neuroscience could provide revolutionary answers to these two closely related problems. Indeed, the human brain has the remarkable property of being able to react at any time, while consuming a reasonable amount of energy, about tens of watts. This system is the result of millions of years of natural selection, and a striking difference between biological neural networks and artificial ones is the representation they use. CNNs, for example, represent information passing from one layer to another as tensors representing visual information densely over the visual topography, while representing different properties in different channels. These networks have been known to mimic many properties of biological systems~\citep{schrimpf_brain-score_2020}, yet, this score does not count rapidity or energy. On the other hand, information in a large majority of biological neural networks is represented as spikes, i.e. prototypical all-or-nothing events whose only parameters are their timing and the address of the neuron that emitted this spike~\citep{paugam-moisy_computing_2012}. Why spikes? It is not yet known why most biological systems have evolved to use this representation, but one major advantage seems to be some form of efficiency. In the human brain, each of the approximately 90 trillion nerve cells emits on average one spike per second which fan out to an average of 10000 synapses. Importantly, each neuron fires with great variability, and up to 300 spikes per second. Many attempts have been made to build artificial spiking neural networks (SNNs), and in particular to achieve the same performance as observed in classical CNNs with a lower energy budget. However, none of these networks has yet surpassed the state-of-the-art performances. Here, we will try to focus on one core property of spiking neurons, inspired by neurobiology to progress in this direction.

@@

Let us first review some existing SNN models. A first motivation for these models was to understand biological observations at the different scales of the central nervous system. For example, the Hodgkin-Huxley model describes the dynamics of the membrane potential of a single nerve cell and specifically explains the role of the different voltage-dependent ion channels. It also explains how the model can produce a spike. Larger scale models introduce an simpler dynamics and an explicit spiking mechanism, for example in the Integrate-and-Fire model. It predicts the emergent states of a recurrently connected neural network~\citep{brunel_phase_2000}. Several models have increased the scale of the system to include more neurons, such as The Virtual Brain (TVB)~\citep{sanz_leon_virtual_2013}, which mixes different scales and includes SNNs with (analog) neural mass models. However, these methods are mainly descriptive with applications in neuroscience but few spiking neural models have had an application in computer vision. Another class of approaches have designed normative models of SNNs that aim to have applications in machine learning. One of these is the SpikeNet algorithm, which uses a purely temporal approach by encoding information using one spike per neuron~\citep{delorme_spikenet_1999}. Another type of SNN using precise spike timing attempts to determine the structure of the network in order to minimize a cost function. This was implemented in the SpikeProp algorithm~\citep{bohte_error-backpropagation_2002} and has been extanded in novel gradient-based methods. The surrogate gradient method is now widely used in methods that attempt to transfer performance from CNNs to SNNs~\citep{zenke_remarkable_2021}. However, the performance of SNNs is still lagging that of analog networks and the question of the advantage of using spikes in machine learning and computer vision remains open.

@@

Let us highlight some innovations in the contributions presented in this paper. First, the generic heterogeneous model is formalized from first principles for optimal detection of the event-based spatio-temporal motifs, whereas~\citet{ghosh_spatiotemporal_2019,yu_stsc-snn_2022} use a correlation-based heuristic, which we have observed to be less efficient. Moreover, in comparison to HOTS~\citep{lagorce_hots_2017} the weights are explainable as they directly inform on the logit (inverse sigmoid of the probability) of detecting each spatio-temporal spiking motif. Another novelty is that the model simultaneously learns the weights and the delays, while for example the polychronization model~\citep{izhikevich_polychronization_2006} only learns the weights using STDP while the delays are randomly drawn and their values frozen. Moreover, the model is evaluated on a realistic task, while models such as the tempotron are tested on simplified problems~\citep{gutig_tempotron_2006}. Another main contribution is to provide a model that is suitable for learning any type of spatio-temporal spiking motifs and that can be trained in a supervised way by providing a dataset of supervision pairs. This allows for a more flexible definition of the model using this properly labeled dataset instead of relying on a careful description of the physical rules governing a task, e.g., the luminance conservation principle for motion detection~\citep{benosman_asynchronous_2012, dardelet_event-by-event_2021}.
%
\subsection{Main limits}
%%%-----------------------------------------------------------------
We have identified several limits to our model, which we will now detail. First, the complete framework is based on a discrete binning of time that is incompatible with the continuous nature of biological time. We have used this binning to be able to efficiently implement the framework on conventional hardware, especially GPUs, and in particular to be able to use three-dimensional convolutions. We have tested the effect of the size of the time bin and shown that it has essentially no impact on the results presented in this paper. This is consistent with the relative robustness of other event-based frameworks such as HOTS~\citep{lagorce_hots_2017}, where the accuracy was not affected when the input spikes were subjected to noisy perturbations up to $1~\ms$~\citep{grimaldi_robust_2022}. This suggests the possibility of analytically including a precision term in the temporal value of the input spikes, a mechanism potentially implemented by the filtering that is implemented by the synaptic time constant of about $5~\ms$. Furthermore, it is possible to circonvent the necessity of using a time discretization by using a purely event-based scheme. Indeed, it is not necessary to compute voltage traces between two spikes~\citep{hanuschkin_general_2010} and it is thus possible to define a purely event-based framework. Such an architecture could provide promising speed gains for the calculations.  \note{pourrait parler du poster bernstein? de nadafian?}

Another restriction is that this model is purely feed-forward. Therefore, the spikes generated by the postsynaptic neurons are produced solely on the basis of information contained in the classical receptive field. However, it is well known that neurons in a layer can interact using lateral interactions, for instance in V1 and that this can be the basis for computational principles~\citep{chavane_revisiting_2022}. For example, the combination of neighboring orientations may contribute to image categorization~\citep{perrinet_edge_2015}. Furthermore, neural information is modulated by feedback information, for example to distinguish a figure from its background~\citep{roelfsema_early_2016} and it has been shown that feedback may be essential for building realistic models of primary visual areas~\citep{boutin_sparse_2020, boutin_effect_2020}, Notably to explain non-linear mechanisms~\citep{boutin_pooling_2022}. It is currently not possible to implement these recurrent connections in our implementation (lateral or feedback), mainly due to our use of convolutions. However, the generic theoretical model is able to include them by inserting new spikes into the list of spikes that are reaching pre-synaptic addresses. While this is possible in theory, it needs to be properly adapted in practice so that these recurrent connections do not amplify neuronal activity outside a homeostatic state (through an extinction or explosion of activity).  
\note{ dire que la prédiction permet de faire de la prediction et que ça explique les résultats de leBec}%~\citep{perrinet_motion-based_2012}

Such recurrent activity would be essential for implementing predictive or anticipatory processes. This is essential in a neural system, as it contains multiple different delays that require temporal alignment~\citep{hogendoorn_predictive_2019}. This has been previously modeled to explain for example the flash-lag illusion~\citep{khoei_flash-lag_2017}. \note{Say that this could be implemented by properly setting the timing of the label} As previously stated, this could be implemented using generalized coordinates (that is variables such as position which are complemented by speed, acceleration, jerk, ...) and ``neurobiologically, the application of delay operators just means changing synaptic connection strengths to take different mixtures of generalized sensations and their prediction errors''~\citep{perrinet_active_2014}. Our proposed model using heterogeneous delays provides an alternate and elegant implementation solution to this problem.
%
\subsection{Perspectives}
%%%-----------------------------------------------------------------
\note{la tache ne met pas en avant les capacites des motifs hetero synaptiques =trop simple et peut se faire en HOTS}
In the definition of our task, we highlighted how the generation of events depends on the spatial gradient in each image. This gradient is defined in both horizontal and vertical dimensions and its maxima are generally oriented. Taken together, these oriented edges form the contours of visual objects in the scene~\citep{koenderink_representation_1987}. Thus, there is a dependence within the event stream between motion and orientation information. It would be crucial to study this interdependence further. This could be initiated by training the model on a dataset with labels providing local orientation. \note{we can say we tried and that it works well}. Studying this dependence will allow us to dissociate these two forms of visual information and to enable us to integrate them. In particular, it will allow us to consider that the definition of motion is more accurate perpendicular to an oriented contour (aka, the aperture problem), thus allowing us to implement recurrent predictive rules, such as those identified to dissociate this problem~\citep{perrinet_motion-based_2012}.
